---
title: "R Programming Week 1: Basic Data Types and Subsetting"
author: "NataliaSmirnova"
date: "Saturday, November 12, 2016"
output: html_document
---

1. Nuts and bolts

Hotkeys: Help -> Keyboard Shortcut Quick Refernce


Working directory:


 - Session - > to set working directory (WD) 
 - getwd() - to check WD
 - dir() - to list files from the working directory
 - ls() - to list names of the objects in the specified environment
 - source ("file name.r")  - to upload and to execute R code


Expressions:

 -  x <- 1
 -  x <- 1:20
 -  x <- c(1:20)

 ## - comment
 
2. Data types:

 - character
 - numbers (real)
 - integer
 - complex
 - logical
 
   Objects: 
 
 - vector (objects of the same class)
 - list (objects of different class)
 - factor
 - matrix
 - data frame



By default all numbers are real, to get integer  - need to use L postfix: e.g. 1L

Special numbers:

 - Inf - infinity
 - NaN - undefined number



Object's attributes

 - name (for all objects exept for matrices), dimnames (for matrices)
 - dimensions (e.g. matrices, arrays, data frames)
 - class (for all objects)
 - length (for all objects)
 - other attributes

Attributes of an object can be assessed with the help of attributes() function. Will return NULL if no attributes set up


2.1 Vectors


functions:

 - c () to concatenate elements into vector
 - vector(class,length)  to create an empty vector



```{r}
x <- c(1,5,7)
y <- c("abc","cde")
z <- c(4+1i,5+1i)


w <- c("r",1)  ## number will be converted to a character
w <- c(TRUE,1) ## logical value will be converted to a number
```

as.data_type  explicit coercion, not always applicable

```{r}
as.numeric(w)

as.logical(w)

as.character(w)

```


2.2 Lists

```{r}
l <- list (1,"a",c(TRUE,FALSE))
names(l)<-c("el_1","el_2","el_3")
class(l)
attributes(l)
print(l)
```


2.3 Matrices

function matrix (nrow,ncol)

```{r}

m<-matrix (nrow=2,ncol=3)

m

attributes (m)

dim (m)

m<- matrix (1:6,2,3)
print (m)
m<-c(1:6)
print(m)
dim(m)<-c(2,3)
print(m)

x<-c(1,2,3)
y<-c(4,5,6)
n<-cbind(x,y)
n<-rbind(x,y)
dimnames(n)<-list(c("r1","r2"),c("c1","c2","c3"))
print(n)
```

2.4 Factors

 - ordered and unordered
 - used to represent categorical values
 
```{r}
f<-factor(c("yes","no","yes","yes","no"),level=c("yes","no")) ## argument"level" sets the order of the levels
print(f)
attributes(f)
names(f)
table(f) ##frequency of values per level
```



2.5 Missing Values

NA OR NAN (undefined mathematical operation)

```{r}
k<-c(1,2,3,NA,4,5,6,NaN,8)
is.na(k)
is.nan(k)
```

2.6 Data Frames

 - stores tabular data (structured tables)
 - special type of list where each element (column of a table) has the same length (number of rows in a table)
 - can store different type of elements in each column
 
```{r}
df1<-data.frame("a",1,TRUE)
df1

row.names(df1)<-"row#1"
df1

colnames(df1)<-c("col#1","col#2","col#3")
df1

df2<-data.frame(1,2,3,4,5,6)
dfm<-data.matrix(df2)
dim(dfm)<-c(2,3)
dfm


df3<-data.frame(col_1 = 1:4, col_2 = c(T,F,T,T))
df3

attributes(df3)
nrow(df3)
ncol(df3)
names(df3)
```

3. Reading/writing data

3.1 Tabular data

 - read.table(file,header,sep,colClasses,nrows,comment.char, skip,StringAsFactors)
 - read.csv(the same parameters exept for 'sep' that shouldn't be defined as it is always ",")
 - write.table()
 
    Parameters:
 
 - file - name of the file/connection (mandatory)
 - header - logical indicating if file has headers (FALSE by default, read.csv function default value is TRue)
 - colClasses - character vector indicating class of each column (length of the vector = to number of columns, if not defined than it takes time for R to define which is important for large data sets)
 - nrows - number of rows in dataset 
 - comment.char - a character string indicating the comment character (everything after the symbol is ignored, # - by default)
 - skip - the number of lines to skip from the beginning of the file
 - StringAsFactors - should character variables be coded as vectors (True by default)
 
 The more parameters defined the faster executes code and less RAM required.
 
 Hints: 
 
  - read first line and define type of each column using class() function. Nrow parameter helps to reduce consumed RAM
  - calculate RAM required  = (number of rows x number of colums x 8 bytes per numeric) / 2^20/1024 = X Gb
 
 
3.2 Text files

 - readLines()
 - writeLines()
 
3.3 Files that include metadata

 - dget () - read - already contains metadata - can be used on a single R object
 - dump/dput () - write -  already contains metadata - can be used on a multiple R objects
 
 Files containing metada are editable but not stable. Therefore, it is recommended to store large data in a textual formats (for vesion control progamms as well). The drawback of a textual format - not space-efficient, often need to be compressed
 
 
3.4 R code
 
 - source()  - can read files created by dump/dget()
 
3.5 Connection interface 

 - file(discription,open) - opens a connection to file
 - gzfile - read file compressed with gzip - allows to read a part of the file
 - bzfile - read file compressed with bzip2 - allows to read a part of the file
 - url - opens connection to the web


    Parameters: 
    - discription - name of the file
    - open ("r" - read only, "w" - write (new file), "a" - appending; "rb", "wb", "ab" - in a binary mode)
    
    
4. Subsetting

4.1. Basic subsetting 

Operators:

 - [ always return an object of the same class as the original
 - [[ is used to extract element of a list or a data frame - type of returned object can be different from list or data frame
 - $ the same as [[ but using object's names
 
```{r}

x<-c(a=1,b=9,c=5,d=7,e=0)
x[1]
x[[1]]
x[3:4]
x[x>6]
u<-x>4
u
```

4.2 Subsetting lists

```{r}

l<-list(num=c(1:10),char=c("a","b","c"),log=c(TRUE,FALSE))
l[1]
l[[1]]
l["char"]
l[["char"]]
l$char
l[c(1,3)]
name<-"log"
l[[name]]
l$name
l[[c(1,5)]]
l[[3]][[1]]
```

4.3 Subsetting matrices

```{r}
m<-matrix(1:10,2,5)
m
m[2,3]
m[1,]
m[,4]
```
[] operator returns not a matrix but rather an element or vector

```{r}
class(m[,4])
```

to preserve the dimension of an object and get a matrix

```{r}
class(m[,4,drop=FALSE])
m[,4,drop=FALSE]
```

4.4 Partial subsetting

```{r}
l[["ch",exact=FALSE]]
l$nu
```


4.5 Removing NA values

```{r}
x<-c(1,2,NA,4,5,NaN,7,8)
bad<-is.na(x)
x[!bad]
```


```{r}
x<-c(1,2,NA,4,5,NaN,7,NaN)
y<-c("a",NA,"c","d",NaN,"f","g",NA)
good<-complete.cases(x,y) #not sensitive to NaN in character vector
good
x[good]
y[good]
```

5. Vectorized Operations

Vectors can added, subtracted, multiplied, divided, compare element-wise.


```{r}
x<-c(2:5)
y<-c(4:7)
x*y
x-y
x+y
x/y
x>=3
y==6
x!=4
```
The same  operators are applicable for matrices exept for multiplication and division:

 - '*'   - element-wise multiplication
 - '%*%' - true matrix multiplication
 - '/' - element-wise division
 - '%*%' - true matricx division
 
